Generating Random Logic Programs Using Constraint Programming

### 2019-12-31 paper/paper.tex
Some examples

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$.

@@

\section{TODO}

\begin{itemize}
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item A constraint for logical equivalence.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

\section{Parameters}

Parameters:
\begin{itemize}
\item maximum number of solutions,
\item $\maxNumNodes{}$: maximum number of nodes in the tree representation of a
  clause,
\item $\maxNumClauses{}$: maximum number of clauses in a program,
\item option to forbid all cycles or just negative cycles,
\item a list of probabilities that are randomly assigned to clauses,
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$,
  \begin{itemize}
  \item maximum arity $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
\end{itemize}

@@

\section{General Constraints}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: might need to include clause heads as well.

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{treeStructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{treeValues}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}
% TODO: examples

Auxiliary variables: $\variable{numNodes}, \variable{numTrees} \in \{ 1, \dots,
\maxNumNodes{} \}$.

\subsection{Nodes}
%, where $\Box$ means the position is either
%  reserved for a variable, or disabled

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{}}$. The node's
  \variable{arity} is a number in $[0, \maxArity{}]$ defined by a
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0.
  \]
\end{constraint}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{itemize}
  \item has 0 children $\iff$ $\variable{treeValues}[i].\variable{name} \in
    \predicates{}$;
  \item has 1 child $\iff$ $\variable{treeValues}[i].\variable{name} = \neg$;
  \item has $>1$ child $\iff$ $\variable{treeValues}[i].\variable{name} \in
    \{ \land, \lor \}$.
  \end{itemize}
\end{constraint}

@@

@@

Our definition of a head of a clause is more restrictive than \cref{def:node}.

\begin{definition} % mention: variables are always ordered, constants are not.
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item $\arrayd{variables}{|\variables{}|}{[0, \maxArity{}]}$: how many times
    each variable appears in the head atom.
  \item $\arrayd{constants}{\maxArity{}}{\constants{} \cup \{ \Box \}}$, where
    $\Box$ denotes that the position is reserved for a variable.
  \end{itemize}
  We also define the \variable{predicate}'s \variable{arity} using the same
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

Parameters that affect this:
\begin{itemize}
\item $\arities{}$ and $(p_a)_{a \in \arities{}}$. Consider even zero arity.
\item $|\predicates{}| = \sum_{a \in \arities{}} p_a$
\item $|\variables{}|$
\item $|\constants{}|$. Consider no variables and no constants.
\item $\maxNumNodes{} \ge 1$
\item $\maxNumClauses{} \ge |\predicates{}|$
\end{itemize}
% TODO: this is without any kind of cycle detection and without probabilities
% describe the parameters of the test cases

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. All constraint variables in the model are integer
variables, but, e.g., if the integer $i$ refers to a logical variable $X$, we
will use $i$ and $X$ interchangeably.

@@

\subsection{TODO}

\begin{itemize}
\item Heads are too restrictive: $\variable{P}(X, Y, X)$ should be allowed.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item A constraint for logical equivalence.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

\subsection{Parameters}

Parameters:
\begin{itemize}
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$ (including zero),
  \begin{itemize}
  \item maximum arity $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
  \begin{itemize}
  \item Each of them can be empty, but $|\constants{}| + |\variables{}| > 0$.
  \end{itemize}
\item a list of probabilities that are randomly assigned to clauses,
\item option to forbid all cycles or just negative cycles,
\item $\maxNumNodes{} \ge 1$: maximum number of nodes in the tree representation
  of a clause,
\item $\maxNumClauses{} \ge |\predicates{}|$: maximum number of clauses in a
  program,
\item maximum number of solutions,
\end{itemize}

@@

Our definition is slightly more involved because we want to impose some
additional constraints, namely, that the variables in the atom are consecutive

\begin{definition} % mention: variables are always ordered, constants are not.
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item $\arrayd{variables}{|\variables{}|}{[0, \maxArity{}]}$: how many times
    each variable appears in the head atom.
  \item $\arrayd{constants}{\maxArity{}}{\constants{} \cup \{ \Box \}}$, where
    $\Box$ denotes that the position is reserved for a variable.
  \end{itemize}
  We define the \variable{predicate}'s \variable{arity} using the
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{treeStructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{treeValues}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

Auxiliary variables: $\variable{numNodes}, \variable{numTrees} \in \{ 1, \dots,
\maxNumNodes{} \}$. The former counts the number of nodes in the main tree. The
latter counts the number of trees in total.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (0, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1, -1) {$\neg$};
    \node (P) at (-1, -2) {$\variable{P}(X)$};
    \node[draw,circle,gray,text=black] (and) at (1, -1) {$\land$};
    \node (Q) at (0, -2) {$\variable{Q}(X)$};
    \node (R) at (2, -2) {$\variable{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \label{fig:example_tree}
  \caption{A tree representation of the formula from \cref{example:formula}}
\end{figure}

\begin{example} \label{example:formula} % TODO: describe (and refer to the figure)
  Let $\maxNumNodes{} = 8$. Then $\neg\variable{P}(X) \lor (\variable{Q}(X)
  \land \variable{P}(X))$...
  \begin{alignat*}{8}
    \variable{treeStructure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{treeValues} &= [{\lor}, &&{\neg}, &&{\land}, \variable{P}(&&X), \variable{Q}(&&X), \variable{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
\end{example}

\subsection{Nodes}

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{}}$. The node's
  $\variable{arity} \in [0, \maxArity{}]$ is defined by a \variable{table}
  constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0\footnote{Zero
      here represents the first variable or constant, regardless of what it is.}.
  \]
\end{constraint}

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\variable{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\variable{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \variable{P},\\
    \variable{arguments} &= [X, 0],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
  $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$ with
  position $i$ skipped.
\end{constraint}
Each constraint corresponds to node $i$ having no children, one child, and
multiple children, respectively.

@@

\section{Managing Clauses}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.

@@

@@

This is without any kind of cycle detection and without probabilities.
% TODO: describe what parameter values were used to test the model.
